RDF Validation of Metadata \\ on Highly-Complex Person-Level Data

### 2015-02-06 paper/paper7/paper7-appendix.tex
disco validation rules

\begin{itemize}
	\item \textbf{{\em DISCO-C-XX:}} 
All {\em disco:Universe}s must also be {\em skos:Concept}s:

\begin{DL}
Universe $\sqsubseteq$ Concept
\end{DL}
\end{itemize}

@@

\begin{itemize}
	\item \textbf{{\em DISCO-C-XX:}}
All {\em sio:SIO\_000367} resources must also be {\em disco:Variable}s:

\begin{DL}
Variable $\equiv$ SIO\_000367
\end{DL}

The Semanticscience Integrated Ontology (SIO)\footnote{https://code.google.com/p/semanticscience/wiki/SIO} provides a simple, integrated ontology of types and relations for rich description of objects, processes and their attributes.
{\em sio:SIO\_000367} is a variable defined as a value that may change within the scope of a given problem or set of operations.
Thus, {\em sio:SIO\_000367} is equivalent to {\em disco:Variable}.
\end{itemize}

@@

\begin{itemize}
	\item \textbf{{\em DISCO-C-XX:}}
If an individual \emph{x} is connected by {\em disco:fundedBy} to an individual \emph{y}, then \emph{x} is also connected by {\em dcterms:contributor} to \emph{y}.

\begin{DL}
fundedBy $\sqsubseteq$ contributor 
\end{DL}
\end{itemize}

@@

\begin{itemize}
	\item \textbf{{\em DISCO-C-XX:}} 
Only {\em disco:Question}s can have {\em disco:responseDomain} relationships.

\begin{DL}
$\exists$ responseDomain.$\top$ $\sqsubseteq$ Question 
\end{DL}
\end{itemize}

@@

\begin{itemize}
	\item \textbf{{\em DISCO-C-XX:}} 
{\em disco:caseQuantity} relationships can only point to literals of the datatype {\em xsd:nonNegativeInteger}.

\begin{DL}
$\top$ $\sqsubseteq$ $\forall$ caseQuantity.nonNegativeInteger \\
\end{DL}
\end{itemize}

@@

\begin{itemize}
	\item \textbf{{\em DISCO-C-XX:}} 
A {\em disco:Variable} may be based on a {\em disco:RepresentedVariable}.
A {\em disco:RepresentedVariable}, however, cannot be based on a {\em disco:Variable}.
This is a kind of mistake which may occur as a semantically equivalent object property for the other direction may also be possible ({\em disco:basisOf}).

\begin{DL}
$basedOn \sqcap basedOn^{-} \sqsubseteq \bot$ 
\end{DL}
\end{itemize}

@@

\begin{itemize}
  \item \textbf{{\em DISCO-C-XX:}}
Within the Disco context, no individual is connected by the object property {\em instrument} to itself

\begin{DL}
$\top$ $\sqsubseteq$ $\neg$ $\exists  instrument . Self$ 
\end{DL}
\end{itemize}

@@

\begin{itemize}
  \item \textbf{{\em DISCO-C-XX:}}
Within the Disco context, {\em skos:Concept}s cannot be related via the object property {\em skos:boader} to themselves

\begin{DL}
Concept $\sqsubseteq$ $\neg$$\exists$ broader.Self. 
\end{DL}

	\item \textbf{{\em DISCO-C-XX:}}
Within the Disco context, {\em skos:Concept}s cannot be related via the object property {\em skos:narrower} to themselves.

\begin{DL}
Concept $\sqsubseteq$ $\neg$$\exists$ narrower.Self. 
\end{DL}
\end{itemize}

@@

\begin{itemize}
	\item \textbf{{\em DISCO-C-XX:}} 
All \emph{Disco} classes are defined to be pairwise disjoint.
The following DL statements holds for each pair of Disco classes:

\begin{DL}
Study $\sqcap$ Variable $\sqsubseteq$ $\perp$\\
\end{DL}
\end{itemize}

@@

\begin{itemize}
	\item \textbf{{\em DISCO-C-XX:}} The abstract of a study (\emph{disco:purpose}) should have a minimum length (\emph{xsd:minInclusive}) of x. 
\end{itemize}

@@

\begin{itemize}
	\item \textbf{{\em DISCO-C-XX:}} Each \emph{disco:Variable} of a given \emph{disco:LogicalDataSet} must have a given prefix for its variable name (\emph{skos:notation}). 
\end{itemize}

@@

\begin{itemize}
	\item \textbf{{\em DISCO-C-XX:}} 
\end{itemize}

@@

\begin{itemize}
	\item \textbf{{\em DISCO-C-XX:}} 
Only {\em disco:Variable}s or {\em disco:Question}s or {\em disco:RepresentedVariable}s can have {\em disco:concept} relationships to {\em skos:Concept}s.

@@

\begin{itemize}
	\item \textbf{{\em DISCO-C-XX:}} 
There must be at least one {\em disco:universe} relationship from {\em disco:Studies} or {\em disco:StudyGroups} to {\em disco:Universe}.

\begin{DL}
Study $\sqcup$ StudyGroup $\sqsubseteq$ $\exists$ universe.Universe \\
\end{DL}
\end{itemize}

@@

\begin{itemize}
	\item \textbf{{\em DISCO-C-XX:}}
Only {\em disco:LogicalDataSet}s can have {\em disco:aggregation} relationships to {\em qb:DataSet}s.

\begin{DL}
LogicalDataSet $\sqsubseteq$ $\forall$ aggregation.DataSet \\
\end{DL}
\end{itemize}

@@

\begin{itemize}
	\item \textbf{{\em DISCO-C-XX:}}
A {\em disco:Questionnaire} has at least one {\em disco:question} relationship to {\em disco:Question}.

\begin{DL}
Questionnaire $\sqsubseteq$ $\geq$1 question.Question
\end{DL}
\end{itemize}

@@

\begin{itemize}
	\item \textbf{{\em DISCO-C-XX:}}
A {\em disco:Question} has exactly 1 {\em disco:universe} relationship to {\em disco:Universe}.

\begin{DL}
Question $\sqsubseteq$ $\geq$1 universe.Universe $\sqcap$ $\leq$1 universe.Universe \\
\end{DL}
\end{itemize}

@@

\begin{itemize}
	\item \textbf{{\em DISCO-C-XX:}}
Within the context of \emph{Disco}, \emph{skos:Concept}s can have either \emph{skos:definition} (when interpreted as DDI concepts) or \emph{skos:notation} and \emph{skos:prefLabel} properties (when interpreted as DDI codes and categories), but not both.

@@

\begin{itemize}
	\item \textbf{{\em DISCO-C-XX}}.
{\em disco:CategoryStatistics} can only have {\em disco:computationBase} relationships to the values \emph{valid} and \emph{invalid} of the datatype {\em rdf:langString}.

\begin{DL}
CategoryStatistics $\equiv$ $\forall$ computationBase.\{valid,invalid\} $\sqcap$ langString \\
\end{DL}
\end{itemize}

@@

\begin{itemize}
	\item \textbf{{\em DISCO-C-XX:}}
{\em disco:percentage} (domain: {\em disco:CategoryStatistics}) literals must be of the datatype {\em xsd:double} whose range should be restricted to be between 0 and 100 (not expressible in DL).
\end{itemize}

@@

\begin{itemize}
	\item \textbf{{\em DISCO-C-XX:}} for each \emph{rdfs:Resource x}, there can be at most one distinct \emph{rdfs:Resource y} such that \emph{x} is connected by \emph{adms:identifier} to \emph{y}.

\begin{DL}
$(\textsf{funct } \textsf{identifier})$
\end{DL}
\end{itemize}

@@

\begin{itemize}
	\item \textbf{{\em DISCO-C-XX:}} 
Only {\em disco:Question}s can have {\em disco:questionText} relationships to literals of the datatype {\em rdf:langString}.

\begin{DL}
$\neg$Question $\sqsubseteq$ $\neg\exists$ questionText.langString
\end{DL}
\end{itemize}

@@

\begin{itemize}
	\item \textbf{{\em DISCO-C-XX}}:
{\em disco:SummaryStatistics} can only have {\em disco:summaryStatisticType} relationships to {\em skos:Concept}s which must be members of the controlled vocabulary {\em ddicv:SummaryStatisticType} which is a {\em skos:ConceptScheme}.

@@

\begin{itemize}
	\item \textbf{{\em DISCO-C-XX}}: \emph{disco:Study} resources must match a given IRI pattern.
\end{itemize}

@@

\begin{itemize}
	\item \textbf{{\em DISCO-C-XX}}:
{\em disco:startDate}s must be before (‘\textless’) {\em disco:endDate}s.
To validate this constraint we bind the variables as follows (P1: {\em disco:startDate}, P2: {\em disco:endDate}, OP: \textless). 
\end{itemize}

@@

The validation of instances data (direct or indirect) exploits the sub-class or sub-property link in a given ontology.
This validation can indicate when the data is verbose (redundant) or expressed at a too general level, and could be improved.
If dcterms:date and one of its sub-properties dcterms:created or dcterms:issued are present, check that the value in dcterms:date is not redundant with dcterms:created or dcterms:issued for ingestion.

\begin{itemize}
	\item \textbf{{\em DISCO-C-USE-SUB-SUPER-RELATIONS-IN-VALIDATION-01}}: 
	If one or more \emph{dcterms:coverage} properties are present, suggest the use of one of its sub-properties \emph{dcterms:spatial} or \emph{dcterms:temporal}.
\end{itemize}

@@

@@

\subsection{Hierarchies}

Hierarchies of DDI concepts

\begin{itemize}
	\item root?
\end{itemize}

>>>>

\begin{itemize}
	\item \textbf{{\em DISCO-C-SUBSUMPTION-01:}} 
All {\em disco:Universe}s must also be {\em skos:Concept}s (\ms{Universe $\sqsubseteq$ Concept}).
\end{itemize}

@@

\begin{itemize}
	\item \textbf{{\em DISCO-C-CLASS-EQUIVALENCE-01:}}
All {\em sio:SIO\_000367} resources must also be {\em disco:Variable}s (\ms{Variable $\equiv$ SIO\_000367}).
The Semanticscience Integrated Ontology (SIO)\footnote{https://code.google.com/p/semanticscience/wiki/SIO} provides a simple, integrated ontology of types and relations for rich description of objects, processes and their attributes.
{\em sio:SIO\_000367} is a variable defined as a value that may change within the scope of a given problem or set of operations.
Thus, {\em sio:SIO\_000367} is equivalent to {\em disco:Variable}.
\end{itemize}

@@

\begin{itemize}
	\item \textbf{{\em DISCO-C-SUB-PROPERTIES-01:}}
If an individual \emph{x} is connected by {\em disco:fundedBy} to an individual \emph{y}, then \emph{x} is also connected by {\em dcterms:contributor} to \emph{y} (\ms{fundedBy $\sqsubseteq$ contributor}). 
\end{itemize}

@@

\begin{itemize}
	\item \textbf{{\em DISCO-C-PROPERTY-DOMAIN-01:}} 
Only {\em disco:Question}s can have {\em disco:responseDomain} relationships (\ms{$\exists$ responseDomain.$\top$ $\sqsubseteq$ Question}).
\end{itemize}

@@

\begin{itemize}
	\item \textbf{{\em DISCO-C-PROPERTY-RANGES-01:}} 
{\em disco:caseQuantity} relationships can only point to literals of the datatype {\em xsd:nonNegativeInteger} (\ms{$\top$ $\sqsubseteq$ $\forall$ caseQuantity.nonNegativeInteger}).
\end{itemize}

@@

\begin{itemize}
	\item \textbf{{\em DISCO-C-ASYMMETRIC-OBJECT-PROPERTIES-01:}} 
A {\em disco:Variable} may be based on a {\em disco:RepresentedVariable}.
A {\em disco:RepresentedVariable}, however, cannot be based on a {\em disco:Variable}.
This is a kind of mistake which may occur as a semantically equivalent object property for the other direction may also be possible ({\em disco:basisOf}) (\ms{$basedOn \sqcap basedOn^{-} \sqsubseteq \bot$}).
\end{itemize}

@@

\begin{itemize}
  \item \textbf{{\em DISCO-C-IRREFLEXIVE-OBJECT-PROPERTIES-01:}}
Within the Disco context, no individual is connected by the object property {\em instrument} to itself (\ms{$\top$ $\sqsubseteq$ $\neg$ $\exists  instrument . Self$}).
\end{itemize}

@@

\begin{itemize}
  \item \textbf{{\em DISCO-C-CLASS-SPECIFIC-IRREFLEXIVE-OBJECT-PROPERTIES-01:}}
Within the Disco context, {\em skos:Concept}s cannot be related via the object property {\em skos:boader} to themselves (\ms{Concept $\sqsubseteq$ $\neg$$\exists$ broader.Self.}). 
	\item \textbf{{\em DISCO-C-CLASS-SPECIFIC-IRREFLEXIVE-OBJECT-PROPERTIES-02:}}
Within the Disco context, {\em skos:Concept}s cannot be related via the object property {\em skos:narrower} to themselves (\ms{Concept $\sqsubseteq$ $\neg$$\exists$ narrower.Self.}). 
\end{itemize}

@@

\begin{itemize}
	\item \textbf{{\em DISCO-C-DISJOINT-CLASSES-01:}} 
All \emph{Disco} classes are defined to be pairwise disjoint.
The following DL statements holds for each pair of Disco classes (\ms{Study $\sqcap$ Variable $\sqsubseteq$ $\perp$}).
\end{itemize}

@@

\begin{itemize}
	\item \textbf{{\em DISCO-C-DATA-PROPERTY-FACETS-01:}} The abstract of a study (\emph{disco:purpose}) should have a minimum length (\emph{xsd:minInclusive}) of x. 
\end{itemize}

@@

\begin{itemize}
	\item \textbf{{\em DISCO-C-LITERAL-PATTERN-MATCHING-01:}} Each \emph{disco:Variable} of a given \emph{disco:LogicalDataSet} must have a given prefix for its variable name (\emph{skos:notation}). 
\end{itemize}

@@

\begin{itemize}
	\item \textbf{{\em DISCO-C-NEGATIVE-LITERAL-PATTERN-MATCHING-01:}} 
\end{itemize}

@@

\begin{itemize}
	\item \textbf{{\em DISCO-C-DISJUNCTION-01:}} 
Only {\em disco:Variable}s or {\em disco:Question}s or {\em disco:RepresentedVariable}s can have {\em disco:concept} relationships to {\em skos:Concept}s.

@@

\begin{itemize}
	\item \textbf{{\em DISCO-C-EXISTENTIAL-QUANTIFICATIONS-01:}} 
There must be at least one {\em disco:universe} relationship from {\em disco:Studies} or {\em disco:StudyGroups} to {\em disco:Universe} (\ms{Study $\sqcup$ StudyGroup $\sqsubseteq$ $\exists$ universe.Universe}).
\end{itemize}

@@

\begin{itemize}
	\item \textbf{{\em DISCO-C-UNIVERSAL-QUANTIFICATIONS-01:}}
Only {\em disco:LogicalDataSet}s can have {\em disco:aggregation} relationships to {\em qb:DataSet}s (\ms{LogicalDataSet $\sqsubseteq$ $\forall$ aggregation.DataSet}).
\end{itemize}

@@

\begin{itemize}
	\item \textbf{{\em DISCO-C-MINIMUM-QUALIFIED-CARDINALITY-RESTRICTIONS-01:}}
A {\em disco:Questionnaire} has at least one {\em disco:question} relationship to {\em disco:Question} (\ms{Questionnaire $\sqsubseteq$ $\geq$1 question.Question}).
\end{itemize}

@@

\begin{itemize}
	\item \textbf{{\em DISCO-C-EXACT-QUALIFIED-CARDINALITY-RESTRICTIONS-01:}}
A {\em disco:Question} has exactly 1 {\em disco:universe} relationship to {\em disco:Universe} (\ms{Question $\sqsubseteq$ $\geq$1 universe.Universe $\sqcap$ $\leq$1 universe.Universe}).
\end{itemize}

@@

\begin{itemize}
	\item \textbf{{\em DISCO-C-CONTEXT-SPECIFIC-EXCLUSIVE-OR-OF-PROPERTY-GROUPS-01:}}
Within the context of \emph{Disco}, \emph{skos:Concept}s can have either \emph{skos:definition} (when interpreted as DDI concepts) or \emph{skos:notation} and \emph{skos:prefLabel} properties (when interpreted as DDI codes and categories), but not both.

@@

\begin{itemize}
	\item \textbf{{\em DISCO-C-ALLOWED-VALUES-01}}.
{\em disco:CategoryStatistics} can only have {\em disco:computationBase} relationships to the values \emph{valid} and \emph{invalid} of the datatype {\em rdf:langString} (\ms{CategoryStatistics $\equiv$ $\forall$ computationBase.\{valid,invalid\} $\sqcap$ langString}).
\end{itemize}

@@

\begin{itemize}
	\item \textbf{{\em DISCO-C-LITERAL-RANGES-01:}}
{\em disco:percentage} (domain: {\em disco:CategoryStatistics}) literals must be of the datatype {\em xsd:double} whose range should be restricted to be between 0 and 100.
\end{itemize}

@@

\begin{itemize}
	\item \textbf{{\em DISCO-C-FUNCTIONAL-PROPERTIES-01:}} for each \emph{rdfs:Resource x}, there can be at most one distinct \emph{rdfs:Resource y} such that \emph{x} is connected by \emph{adms:identifier} to \emph{y} (\ms{$(\textsf{funct } \textsf{identifier})$}).
\end{itemize}

@@

\begin{itemize}
	\item \textbf{{\em DISCO-C-CLASS-SPECIFIC-PROPERTY-RANGE-01:}} 
Only {\em disco:Question}s can have {\em disco:questionText} relationships to literals of the datatype {\em rdf:langString} (\ms{$\neg$Question $\sqsubseteq$ $\neg\exists$ questionText.langString}).
\end{itemize}

@@

\begin{itemize}
	\item \textbf{{\em DISCO-C-MEMBERSHIP-IN-CONTROLLED-VOCABULARIES-01}}:
{\em disco:SummaryStatistics} can only have {\em disco:summaryStatisticType} relationships to {\em skos:Concept}s which must be members of the controlled vocabulary {\em ddicv:SummaryStatisticType} which is a {\em skos:ConceptScheme}.

@@

\begin{itemize}
	\item \textbf{{\em DISCO-C-IRI-PATTERN-MATCHING}}: \emph{disco:Study} resources must match a given IRI pattern.
\end{itemize}

@@

\begin{itemize}
	\item \textbf{{\em DISCO-C-LITERAL-VALUE-COMPARISON-01}}:
{\em disco:startDate}s must be before (‘\textless’) {\em disco:endDate}s.
To validate this constraint we bind the variables as follows (P1: {\em disco:startDate}, P2: {\em disco:endDate}, OP: \textless). 
\end{itemize}

@@

The validation of instances data (direct or indirect) exploits the sub-class or sub-property link in a given ontology.
This validation can indicate when the data is verbose (redundant) or expressed at a too general level, and could be improved.
If \emph{dcterms:date} and one of its sub-properties \emph{dcterms:created} or \emph{dcterms:issued} are present, e.g., check that the value in \emph{dcterms:date} is not redundant with \emph{dcterms:created} or \emph{dcterms:issued} for ingestion.

\begin{itemize}
	\item \textbf{{\em DISCO-C-USE-SUB-SUPER-RELATIONS-IN-VALIDATION-01}}: 
	If one or more \emph{dcterms:coverage} properties are present, suggest the use of one of its sub-properties \emph{dcterms:spatial} or \emph{dcterms:temporal}.
	\item \textbf{{\em DISCO-C-USE-SUB-SUPER-RELATIONS-IN-VALIDATION-02}}: 
	If the \emph{dcterms:contributor} property is present, suggest the use of one of its sub-properties, e.g. \emph{disco:fundedBy}.
\end{itemize}

@@

In most library applications, cardinality shortcuts tend to appear in pairs, with repeatable/non-repeatable establishing maximum cardinality and optional/mandatory establishing minimum cardinality.
These are shortcuts for more detailed \emph{cardinality restrictions}.

@@

Some constraints require aggregating multiple values, especially via \emph{COUNT}, \emph{MIN} and \emph{MAX}.

\begin{itemize}
	\item \textbf{{\em DISCO-C-AGGREGATION-01}}: calculate the number of theoretical concepts in the thematic classification of a given study.
  \item \textbf{{\em DISCO-C-AGGREGATION-02}}: calculate the number of variables of a data set.
	\item \textbf{{\em DISCO-C-AGGREGATION-03}}: calculate the number of questions in a given questionnaire.
	\item \textbf{{\em DISCO-C-AGGREGATION-04}}: the number of codes of a given variable must be below a maximum value.
	\item \textbf{{\em DISCO-C-AGGREGATION-05}}: the number of questions of a given questionnaire must exactly be a given value.
	\item \textbf{{\em DISCO-C-AGGREGATION-06}}: the cumulative percentage of all codes of a given variable must be 100.
  \item \textbf{{\em DISCO-C-AGGREGATION-07}}: the absolute frequency of all valid codes of a given variable must be equal to a given value.
\end{itemize}

\section{SKOS-Related RDF Validation Requirements / Constraint Types}
\tb{from other paper of Antoine}
\subsection{Hierarchies}

\begin{itemize}
	\item \textbf{{\em DISCO-C-HIERARCHIES-01}}: there must be exactly one root in the hierarchy of DDI concepts. 
\end{itemize}
